A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-22 tidyindex.qmd
remove the notation  of q in dimension reduction

| Section | Module                  | Input                                                                   | Operation                           | Output                                                                        |
|-------|---------------|---------|---------------|--------------------------|
| 3.1   | Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
| 3.2   | Spatial processing      | $x_{tj}(s)$                                                             | $g[x_{tj}(s)]$                      | $x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}$                 |
| 3.3   | Variable transformation | $x_{j}(s; t)$                                                           | $T[x_j(s;t)]$                       | $x^{\text{Trans}}_j(s;t)$                                                     |
| 3.4   | Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
| 3.5   | Dimension reduction     | $\mathbf{x}(s;t)$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q$      |
| 3.6   | Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $P_j(s;t) \quad P(.) \in [0, 1]$                   |
| 3.7   | Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
| 3.8   | Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
| 3.9   | Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                      |

@@

Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$, as the input. It summarises the high-dimensional information into a lower-dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < q$, as the output. The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided by statistical methods. For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) [@SPEI] calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$). This is the only step that differs from the Standardized Precipitation Index (SPI), and can be considered to be a dimension reduction using a particular linear combination.

>>>>

| Section | Module                  | Input                                                                   | Operation                           | Output                                                                        |
|-------|---------------|---------|---------------|--------------------------|
| 3.1   | Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
| 3.2   | Spatial processing      | $x_{tj}(s)$                                                             | $g[x_{tj}(s)]$                      | $x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}$                 |
| 3.3   | Variable transformation | $x_{j}(s; t)$                                                           | $T[x_j(s;t)]$                       | $x^{\text{Trans}}_j(s;t)$                                                     |
| 3.4   | Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
| 3.5   | Dimension reduction     | $\mathbf{x}(s;t)$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p$      |
| 3.6   | Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $P_j(s;t) \quad P(.) \in [0, 1]$                   |
| 3.7   | Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
| 3.8   | Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
| 3.9   | Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                      |

@@

Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$, or a subset of variables in $\mathbf{x}(s;t)$, as the input. It summarises the high-dimensional information into a lower-dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output. The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided by statistical methods. For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) [@SPEI] calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$). This is the only step that differs from the Standardized Precipitation Index (SPI), and can be considered to be a dimension reduction using a particular linear combination.

### 2023-08-22 tidyindex.tex
remove the notation  of q in dimension reduction

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
3.1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
\(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
3.2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
\(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
3.3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
\(x^{\text{Trans}}_j(s;t)\) \\
3.4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
\(x^{\text{Scale}}_j(s;t)\) \\
3.5 & Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\)
& \(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q\) \\
3.6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
\(P_j(s;t) \quad P(.) \in [0, 1]\) \\
3.7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
\(z_j(s; t)\) \\
3.8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
3.9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
\(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\end{longtable}

@@

Dimension reduction takes the multivariate information
\(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\), as the input. It
summarises the high-dimensional information into a lower-dimension
representation \(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\)
and \(d < q\), as the output. The transformation can be based on
domain-specific knowledge, originating from theories describing the
underlying physical processes, or guided by statistical methods. For
example, the Standardized Precipitation-Evapotranspiration Index (SPEI)
(Beguería and Vicente-Serrano 2017) calculates the difference \(D\)
between precipitation (\(P\)) and potential evapotranspiration
(\(\text{PET}\)), using a water balance model (\(D = P - \text{PET}\)).
This is the only step that differs from the Standardized Precipitation
Index (SPI), and can be considered to be a dimension reduction using a
particular linear combination.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Section
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
3.1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
\(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
3.2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
\(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
3.3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
\(x^{\text{Trans}}_j(s;t)\) \\
3.4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
\(x^{\text{Scale}}_j(s;t)\) \\
3.5 & Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\)
& \(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p\) \\
3.6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
\(P_j(s;t) \quad P(.) \in [0, 1]\) \\
3.7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
\(z_j(s; t)\) \\
3.8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
3.9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
\(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

Dimension reduction takes the multivariate information
\(\mathbf{x}(s;t)\), where \(\mathbf{x} \subseteq \mathbb{R}^p\), or a
subset of variables in \(\mathbf{x}(s;t)\), as the input. It summarises
the high-dimensional information into a lower-dimension representation
\(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\) and \(d < p\),
as the output. The transformation can be based on domain-specific
knowledge, originating from theories describing the underlying physical
processes, or guided by statistical methods. For example, the
Standardized Precipitation-Evapotranspiration Index (SPEI) (Beguería and
Vicente-Serrano 2017) calculates the difference \(D\) between
precipitation (\(P\)) and potential evapotranspiration (\(\text{PET}\)),
using a water balance model (\(D = P - \text{PET}\)). This is the only
step that differs from the Standardized Precipitation Index (SPI), and
can be considered to be a dimension reduction using a particular linear
combination.
